---
title: "DATA 621 HW 1"
author: "Jian Quan Chen"
date: "2024-02-22"
output: html_document
---

# Introduction

In this assignment, we are given a baseball training and evaluation dataset. The data spans from 1871 to 2006, each row representing a baseball team's performance that year. The statistics were all adjusted to reflect a 162 game season. Our objective is to construct a multiple linear regression model of the training data to predict the number of wins for a team. 

# 1. Data Exploration

```{r}
library(tidyverse)
library(corrplot)
```

```{r}
df <- read_csv("https://raw.githubusercontent.com/LeJQC/DATA-621-Group-2/main/HW1/moneyball-training-data.csv", show_col_types = FALSE)
```

There are 2,276 rows and 17 columns in the dataset. The response variable is `TARGET_WINS` and the remaining 15 variables, with the exception of the `INDEX` column, are predictor variables. 
```{r}
glimpse(df)
```

```{r}
# Setting index column to index
rownames(df) <- df$INDEX
df$INDEX <- NULL

# Sumarry table
knitr::kable(summary(df))
```

We can see from the summary, that the mean of `TARGET_WINS` is 80.79, which is about half the games in a baseball season. There are also a couple of columns with NA values, specifically `TEAM_BATTING_HBP` has over 2000 missing values. 

Let's look at the distribution of the data.
```{r}
df_long <- df %>%
  pivot_longer(
    cols = everything(), 
    names_to = "variable",
    values_to = "value"
  )

df_long %>%
  ggplot(aes(value)) + 
  geom_density(fill = "blue") + 
  facet_wrap(~variable, scales ="free", ncol = 4) +
  labs(x = element_blank(), y = element_blank())
```

The `TARGET_WINS`, `TEAM_BATTING_2B`, `TEAM_BATTING_HBP`, and `TEAM_FIELDING_DP`  variables show a normal distribution. The `TEAM_BATTING_HR`, `TEAM_BATTING_SO`, and `TEAM_PITCHING_HR` variables show a bimodal distribution.

```{r}
df %>% 
  cor(use = "complete.obs") %>%
  corrplot(method = "color", tl.col = "black")
```

From this visual, wins seem to be most correlated with `TEAM_PITCHING_H`, `TEAM_PTICHING_BB`,`TEAM_BATTING_BB`, `TEAM_BATTING_2B`, and `TEAM_BATTING_H`.

```{r}
cor(df)
```

# 2. Data Preparation
Checking for missing values within the dataset by creating flags for every column
```{r}
# Loop through columns
for (col_name in names(df)) {
  missing <- is.na(df[[col_name]])
  output <- paste(col_name,"missing values?",any(missing))
  print(output)
}
```


Since the `TEAM_BATTING_HBP` variable was missing 2000 values, we will just remove this column from the dataset. The other columns that had missing values (`TEAM_BATTING_SO`, `TEAM_BASERUN_SB`, `TEAM_BASERUN_CS`, `TEAM_PITCHING_SO`, and `TEAM_FIELDING_DP`) will be replaced with the median value of that variable. 

```{r}
df <- df %>% select(-TEAM_BATTING_HBP)

na_variables <- c("TEAM_BATTING_SO", "TEAM_BASERUN_SB", "TEAM_BASERUN_CS", "TEAM_PITCHING_SO", "TEAM_FIELDING_DP")

for (col in na_variables) {
  median_value <- median(df[[col]], na.rm = TRUE)
  df[[col]][is.na(df[[col]])] <- median_value
}

summary(df)
```


Checking for any missing values within the dataset
```{r}
which(is.na(df))
 
# count total missing values 
print("Count of total missing values  ")
sum(is.na(df))
```

Now there is no more missing values within the dataset






